Nuclei Segmentation in Histopathology images using Deep Neural Networks

### 2016-11-07 Paper.tex
correction 1

%% Contribution as a dataset, small introduction
One of the main contributions of this paper is the now public available 
nuclei detection dataset within HE stainned histopathology images which 
can be found at website %\url{www.cbio.ensmp.fr:/pnaylor/download/dataset.zip} 
(not a true address). This annotated dataset provides images clustered by 
patients. Each patient has at least 3 annotated $512 \times 512$ HE 
stainned histopathology images with their associated ground truth. Each 
ground truth image is a $512 \times 512$ where each pixel value above 
$0$ is considered as a pixel belonging to a nuclei. The differences in 
values of these pixels denote different nucleus, such an annotation can 
be used in several processing step which does or doesn't take into 
account clutered nuclei. See figure \ref{fig:annotation} for an example of 
three annotated images. This annotation was conducted via the help of 
the software ITK-snap and were annotated by the authors of the paper.

@@

%%Data set description 
In this data set we have annotated a considerous amount of cells, 
including normal breast cells (epithelial and lobule cells),  cancerous 
breast cells, fibroblasts,  endothelial cells, fat cells, macrophage cells and 
inflammatory cells (lymphocytes and plasmocytes). For the moment, cell types have not been seperated, which means that in this dataset every annotation corresponds to a global type of cell.

@@

\section{Methodology}
\label{sec:method}
%% Segmentation task
Let A be the space of RGB images, A can typically be 
$\mathbb{R}^{n \times p \times 3}$ and let B the space B the space of annotation 
images, in our case $\{0,1\}^{n \times p}$. We have a set of 
$(A_l,B_l)_{l \in [1, N]}$ for a supervised learning approach. Our goal is 
a prediction task 
named as semantic segmentation, we wish to maximize the prediction of 
an unseen element belonging to B given an new element in A. We maximize 
thus prediction by modelizing our prediction function as the softmax output of a deep neural 
network. We find the model parameters by minimizing a log loss function 
defined as:
 $\frac{1}{\sum_{i,j}w_{ij}} \sum_{i,j} \sum_k w_{i,j} t_{i,j,k} \log (\widehat{p_{i,j,k}})$
, where $k$ designates a certain label, $w_{i,j}$ is 
certain weight given to pixel $i,j$, $t_{i,j,k}$ is equal to 1 if pixel $i,j$ is 
of class $k$ and $\widehat{p_{i,j,k}}$ designates the estimated probability 
of pixel $i,j$ of being $k$ via the softmax output of the neural network.
We minimize the loss function via stochastic gradient descent.

%% Validation scheme
We have our training set $(A_l,B_l)_{l \in [1, N]}$ where N is equal to 
34, how ever each element $A_l$ belongs to a certain patients and 
several elements $A_l$ can belong to the same patient. As we are 
dealing with histopathology images, it is know that samples can widely 
vary from one patient to the other. We thus validate our model by a 
leave one patient out scheme. Our validation scheme is as followed, for a 
given set of hype parameters, we 
train our model on every patient except one that is used for validation. 
Our final score is averaged over all patients.
Several metrics assess the quality of the model, the accuracy (Acc), the 
F1 score and a performance (Perf) score which is the mean between the 
true positive rates and the true negative rates.

@@

%% Mathematical Morphology
The results obtained by the Deep Neural Network were encouraging, but
touching nuclei were often segmented as one single object. This is not
surprising, as the error at a pixel level seems very minor, and no
shape prior is used in the model. Solving this issue by weighting the
error term for pixels between objects did not solve the issue for our
case (data not shown). However, we did observe that for touching and
even partially overlapping nuclei, the posterior probability at the
nucleus border is systematically lower than in
the putative center of the nucleus, but may still be larger than
typically applied thresholds.This leads to the following rationale. As in the center of nuclei, the
posterior probability is maximal, we can readily assume that the local
maxima of the posterior correspond to putative nuclei, which we call candidates. Let
$\mathcal{X}=(x_1,x_2,\ldots,x_N)$be a path that joins two candidates
(i.e. $\forall i x_i$ and $x_{i+1}$ are neighbor pixels) and
$\mathcal{P}=(p_1,p_2,\ldots,p_N)$ the corresponding posterior
probabilities. Without loss of generality, we assume $p_1\leq P_N$.
We define for each path $\mathcal{X}$ a cost
$C(\mathcal{P})=\max_{i=2 \ldots N}{p_1-p_i}$, which is the maximal
decrease in posterior probability along the path, when starting from
the candidate with lower probability.  Considering all paths joining two
candidates, we can now state a criterion that allows us to decide whether to
perform a split (and thus accept the candidates as being different
nuclei) or not: if all the paths involve a decrease in probability
larger than a parameter $\lambda$, we will perform a
split. Conversely, if we can find a path that joins the two candidates
with a probability decrease smaller or equal to $\lambda$, no split
will be performed: we argue that in this case, it is probably one
single object. Hence, the split is performed if: 
\begin{equation}
\min_{\mathcal{P}}C(\mathcal{P}) = \min_{\mathcal{P}} \{\max_{i=2
  \ldots N}{p_1-p_i}\} > \lambda
\end{equation}
where $\lambda$ is a free parameter. This is actually nothing else
than the morphological dynamics\cite{Grimaud1992}. The actual split
locations are thus obtained by applying the watershed algorithm,
starting from the local maxima of the posterior probability map with
morphological dynamics larger than a free parameter $\lambda$.

@@

\section{Different architectures/Results}
\label{sec:results}
We experience with 4 known arhitectures in semantic segmentation, one 
will be named BaochuanNet, Fully Convolutionnal Net (FCN), DeconvNet, 
UNet. The most basic architecture, BaochuanNet, consists of 4 
convolutionnal layer where each convolutionnal layer has 8 feature map. 
This net, being not deep, has the advantage of being less 
computationnally intensive. FCN is a first attempt of applying "deep 
feature" representations to the task of semantic segmentation. This 
archicture has the advantage of re-using a classical deep learning 
architecture with added on upsampling layer and skip layers. The 
upsampling layers enable the network to learn a pixel level classification 
and the skip layers enable the network to fuse different levels of 
abstraction to the final prediction. This model can be fine tuned with a 
set of pretrained-weights extracted from the classical deep learning 
architecture, for a more detailed explaination please refer to \cite{long2015fcn}. DeconvNet is also based on a classical architecture, 
however, in this network their are no skip layer as one intends to learn 
the proper upsampling threw repeated deconvolution and convolution 
layers. This model can also be fine tuned, for more information please 
refer to the original paper \cite{noh2015learning}. We did not
ensemble classifier as suggested in 
\cite{noh2015learning} as our FCN and DeconvNet classifier achieve very similar scores.
% The final architecture that we tried is the UNet. This network combines 
%both contribution of the above networks, the UNet has a similar 
%architecture to the DeconvNet with skip layers.
\section{Result and discussion}
\label{sec:result}
We conducted our experiments on GPUs via the caffe framework, \cite{jia2014caffe}. We present our results in table \ref{tab:res} where we averaged the metrics other all left out patients. We also added, for illustration purposes \ref{fig:prediction}, image prediction and associated probability maps to evaluate some differences between the classifiers. We notice that the simpler net, Baochuan Net seems to have learnt simple rules based on color information. As we can not notice by the "holes" in the probability map when the interior of the cell is brighter. On the contrary, deeper nets as FCN and DeconvNet have learnt deeper features and seem capable of recognizing whole cells. The small noticable differences in \ref{fig:prediction} are due to a random crop, this results in a slightly shiffted image for each one of the predictions.
\begin{table}
\begin{tabular}{|c|c|c|c|}
\hline
  & Baochuan & DeconvNet & FCN \\
 \hline
Accuracy  &       0.936 & 0.968 &0.958  \\
IU   &    0.759 &     0.856 & 0.857 \\
Recall     &       0.744  &     0.856 & 0.855 \\
Precision   &       0.741 &     0.858 & 0.878\\
F1    &       0.742 &     0.856 & 0.866 \\
TP       &       0.744 &     0.856 & 0.855  \\
TN      &       0.963 &     0.981 &  0.977\\
Performance    &       0.853 &     0.918 & 0.916\\
Pixel error  &  0.063 &     0.032 & 0.042\\
\hline
\end{tabular}
\caption{\label{tab:res}Results}
\label{tab:res}
\end{table}

%%with Unet
%\begin{tabular}{|c|c|c|c|c|c|}
% & UNet & Baochuan & DeconvNet & FCN & Ensemble \\
%\hrule
%Accuracy &       0.89 &       0.936 & 0.968 &0.958 & \\
%IU   &       0.70 &       0.759 &     0.856 & 0.857 &\\
%Recall    &       0.75 &       0.744  &     0.856 & 0.855 &\\
%Precision    &       0.64 &       0.741 &     0.858 & 0.878&\\
%F1       &       0.69 &       0.742 &     0.856 & 0.866&  \\
%TP             &       0.75 &       0.744 &     0.856 & 0.855 & \\
%TN             &       0.92 &       0.963 &     0.981 &  0.977&\\
%Performance    &       0.84 &       0.853 &     0.918 & 0.916&\\
%Pixel error  &       0.11 &       0.063 &     0.032 & 0.042&\\
%\end{tabular}

@@

\begin{minipage}[b]{0.48\linewidth}
  \centering
  \centerline{\includegraphics[height=4.0cm]{BaochuanB}}
%  \vspace{1.5cm}
  \centerline{BaochuanNet}\medskip
\end{minipage}
\hfill
\begin{minipage}[b]{.48\linewidth}
  \centering
  \centerline{\includegraphics[height=4.0cm]{BaochuanP}}
%  \vspace{1.5cm}
  \centerline{Probability map}\medskip
\end{minipage}

>>>>

%% Contribution as a dataset, small introduction
One of the main contributions of this paper is the now public available 
nuclei detection dataset within HE stainned histopathology images which 
can be found at website %\url{www.cbio.ensmp.fr:/pnaylor/download/dataset.zip} 
(not a true address). This annotated dataset provides images clustered by 
patients. Each patient has at least 3 annotated $512 \times 512$ HE 
stainned histopathology images with their associated ground truth. Each 
ground truth image is a $512 \times 512$ image where each pixel value above 
$0$ is considered as a pixel belonging to a nuclei. The differences in 
values of these pixels denote different nucleus, such an annotation can 
be used in several processing step which does or doesn't take into 
account clutered nuclei. See figure \ref{fig:annotation} for an example of 
three annotated images. This annotation was conducted via the help of 
the software ITK-snap, \cite{py06nimg} http://www.itksnap.org, and were annotated by the authors of the paper.

@@

%%Data set description 
In this data set we have annotated a considerous amount of cells, 
including normal breast cells (epithelial and lobule cells),  cancerous 
breast cells, fibroblasts,  endothelial cells, fat cells, macrophage cells and 
inflammatory cells (lymphocytes and plasmocytes). For the moment, cell types have not been seperated, which means that in this dataset every annotation corresponds to a global type: cell.

@@

\section{Methodology}
\label{sec:method}
%% Segmentation task
Let A be the space of RGB images, A can typically be 
$\mathbb{R}^{n \times p \times 3}$ and let B the space B the space of annotation 
images, in our case $\{0,1\}^{n \times p}$. We have a set of 
$(A_l,B_l)_{l \in [1, N]}$ for a supervised learning approach. Our goal is 
a prediction task 
named as semantic segmentation, we wish to maximize the prediction of 
an unseen element belonging to B given an new element in A. We maximize 
thus prediction by modelizing our prediction function as the softmax output of a deep neural 
network. We find the model parameters by minimizing a log loss function 
defined as:
 $\frac{1}{\sum_{i,j}w_{ij}} \sum_{i,j} \sum_k w_{i,j} t_{i,j,k} \log (\widehat{p_{i,j,k}})$
, where $k$ designates a certain label, $w_{i,j}$ is a 
certain weight given to pixel $i,j$, $t_{i,j,k}$ is equal to 1 if pixel $i,j$ is 
of class $k$ and $\widehat{p_{i,j,k}}$ designates the estimated probability 
of pixel $i,j$ of being $k$ via the softmax output of the neural network.
We minimize the loss function via stochastic gradient descent.

%% Validation scheme
We have our training set $(A_l,B_l)_{l \in [1, N]}$ where N is equal to 
33, how ever each element $A_l$ belongs to a certain patients and 
several elements $A_l$ can belong to the same patient. As we are 
dealing with histopathology images, it is know that samples can widely 
vary from one patient to the other. We thus validate our model by a 
leave one patient out scheme. Our validation scheme is as followed, for a 
given set of hype parameters, we 
train our model on every patient except one that is used for validation. 
Our final score is averaged over all patients.
Several metrics assess the quality of the model, the accuracy, the 
F1 score and a performance score which is the mean between the 
true positive rate and the true negative rate.

@@

%% Mathematical Morphology
The results obtained by the Deep Neural Network were encouraging, but
touching nuclei were often segmented as one single object. This is not
surprising, as the error at a pixel level seems very minor, and no
shape prior is used in the model. Solving this issue by weighting the
error term for pixels between objects did not solve the issue for our
case (data not shown). However, we did observe that for touching and
even partially overlapping nuclei, the posterior probability at the
nucleus border is systematically lower than in
the putative center of the nucleus, but may still be larger than
typically applied thresholds.This leads to the following rationale. As in the center of nuclei, the
posterior probability is maximal, we can readily assume that the local
maxima of the posterior correspond to putative nuclei, which we call candidates. Let
$\mathcal{X}=(x_1,x_2,\ldots,x_N)$ be a path that joins two candidates
(i.e. $\forall i, \ x_i$ and $x_{i+1}$ are neighbor pixels) and
$\mathcal{P}=(p_1,p_2,\ldots,p_N)$ the corresponding posterior
probabilities. Without loss of generality, we assume $p_1\leq P_N$.
We define for each path $\mathcal{X}$ a cost
$C(\mathcal{P})=\max_{i=2 \ldots N}{p_1-p_i}$, which is the maximal
decrease in posterior probability along the path, when starting from
the candidate with lower probability.  Considering all paths joining two
candidates, we can now state a criterion that allows us to decide whether to
perform a split (and thus accept the candidates as being different
nuclei) or not: if all the paths involve a decrease in probability
larger than a parameter $\lambda$, we will perform a
split. Conversely, if we can find a path that joins the two candidates
with a probability decrease smaller or equal to $\lambda$, no split
will be performed: we argue that in this case, it is probably one
single object. Hence, the split is performed if: 
\begin{equation}
\min_{\mathcal{P}}C(\mathcal{P}) = \min_{\mathcal{P}} \{\max_{i=2
  \ldots N}{p_1-p_i}\} > \lambda
\end{equation}
where $\lambda$ is a free parameter. This is actually nothing else
than the morphological dynamics\cite{Grimaud1992}. The actual split
locations are thus obtained by applying the watershed algorithm,
starting from the local maxima of the posterior probability map with
morphological dynamics larger than a free parameter $\lambda$.

@@

\section{Different architectures/Results}
\label{sec:results}
We experience with 3 known arhitectures in semantic segmentation, named PangNet, Fully Convolutionnal Net (FCN) and DeconvNet. The most basic architecture, PangNet, consists of 4 
convolutionnal layer where each convolutionnal layer has 8 feature map, for more information please refer to \cite{pang2010cell}. 
This net, being not deep, has the advantage of being less 
computationnally intensive. FCN is a first attempt of applying "deep 
feature" representations to the task of semantic segmentation. This 
archicture has the advantage of re-using a classical deep learning 
architecture with added on upsampling layer and skip layers. The 
upsampling layers enable the network to learn a pixel level classification 
and the skip layers enable the network to fuse different levels of 
abstraction to the final prediction. This model can be fine tuned with a 
set of pretrained-weights extracted from the classical deep learning 
architecture, for a more detailed explaination please refer to \cite{long2015fcn}. DeconvNet is also based on a classical architecture, 
however, in this network their are no skip layer as one intends to learn 
the proper upsampling threw repeated deconvolution and convolution 
layers. This model can also be fine tuned, for more information please 
refer to the original paper \cite{noh2015learning}. We did not
ensemble classifier as suggested in 
\cite{noh2015learning} as our FCN and DeconvNet classifier achieve very similar scores on the test sets.
% The final architecture that we tried is the UNet. This network combines 
%both contribution of the above networks, the UNet has a similar 
%architecture to the DeconvNet with skip layers.
\section{Result and discussion}
\label{sec:result}
We conducted our experiments on GPUs via the caffe framework, \cite{jia2014caffe}. We present our results in table \ref{tab:res} where we averaged the metrics other all left out patients. We also added, for illustration purposes \ref{fig:prediction}, image prediction and associated probability maps to evaluate some differences between the classifiers. We notice that the simpler net, PangNet seems to have learnt simple rules such as color information. As we can not notice by the "holes" in the probability map when the interior of the cell is brighter. On the contrary, deeper nets as FCN and DeconvNet have learnt deeper features and seem capable of recognizing whole cells. .
\begin{table}
\begin{tabular}{|c|c|c|c|}
\hline
  & PangNet & DeconvNet & FCN \\
 \hline
Accuracy  &       0.936 & 0.968 &0.958  \\
IU   &    0.759 &     0.856 & 0.857 \\
Recall     &       0.744  &     0.856 & 0.855 \\
Precision   &       0.741 &     0.858 & 0.878\\
F1    &       0.742 &     0.856 & 0.866 \\
TP       &       0.744 &     0.856 & 0.855  \\
TN      &       0.963 &     0.981 &  0.977\\
Performance    &       0.853 &     0.918 & 0.916\\
Pixel error  &  0.063 &     0.032 & 0.042\\
\hline
\end{tabular}
\caption{Results}
\label{tab:res}
\end{table}

%%with Unet
%\begin{tabular}{|c|c|c|c|c|c|}
% & UNet & PangNet & DeconvNet & FCN & Ensemble \\
%\hrule
%Accuracy &       0.89 &       0.936 & 0.968 &0.958 & \\
%IU   &       0.70 &       0.759 &     0.856 & 0.857 &\\
%Recall    &       0.75 &       0.744  &     0.856 & 0.855 &\\
%Precision    &       0.64 &       0.741 &     0.858 & 0.878&\\
%F1       &       0.69 &       0.742 &     0.856 & 0.866&  \\
%TP             &       0.75 &       0.744 &     0.856 & 0.855 & \\
%TN             &       0.92 &       0.963 &     0.981 &  0.977&\\
%Performance    &       0.84 &       0.853 &     0.918 & 0.916&\\
%Pixel error  &       0.11 &       0.063 &     0.032 & 0.042&\\
%\end{tabular}

@@

\begin{minipage}[b]{0.48\linewidth}
  \centering
  \centerline{\includegraphics[height=4.0cm]{BaochuanB}}
%  \vspace{1.5cm}
  \centerline{PangNet}\medskip
\end{minipage}
\hfill
\begin{minipage}[b]{.48\linewidth}
  \centering
  \centerline{\includegraphics[height=4.0cm]{BaochuanP}}
%  \vspace{1.5cm}
  \centerline{Probability map}\medskip
\end{minipage}
